\chapter{Architecture}

\section{The logical processor}

A \textbf{Logical Processor} (LP) is one hardware core. The kernel shards its
state per-LP using \texttt{PerLp<T>}: an array of \texttt{RwLock<T>}, one slot
per LP, indexed by the LP's identity. The LP identity is stored in a
register (\texttt{TPIDR\_EL1} on AArch64, \texttt{TSC\_AUX} on x86-64) and
read with a single instruction --- it is always safe to call from any context,
including interrupt handlers.

LP placement is \textbf{not advisory}: an LP \emph{is} a core. When a sitas
shard is bound to an LP it is bound to a core by construction. No
\texttt{sched\_setaffinity} race, no cpuset surprises.

\section{Per-LP state}

\paragraph{PerLp<T>} A boxed slice of \texttt{RwLock<T>}, one per LP. The lock
masks interrupts on acquire, preventing ISR self-deadlock. Used for data
accessed from both thread context and interrupt context (timer queues, LAPIC
registers, interrupt depth counters).

\paragraph{ShardLocal<T>} (newer, sitas-inspired.) A lock-free per-LP container
that stores its slots behind \texttt{UnsafeCell<T>} rather than
\texttt{RwLock<T>}. Access is through a closure (\texttt{with(fn)} or
\texttt{try\_with(fn)}) that asserts:
\begin{itemize}
    \item The caller is on the owning LP (panics otherwise).
    \item The slot is not already borrowed by a re-entrant call (a per-LP
        \texttt{AtomicBool} borrow flag, RAII-guarded).
\end{itemize}
The closure ensures the \texttt{\&mut T} never escapes the accessor. Use
\texttt{PerLp<T>} when interrupt-safety or cross-LP access is needed; use
\texttt{ShardLocal<T>} for single-LP, thread-only data (it has no spin-lock
acquire cost).

\section{The observer / waker model}

CharlotteOS's event mechanism reduces to one pattern: a thread registers its
\texttt{Waker} (an \texttt{Observer} wrapping its \texttt{ThreadId}) on an
\texttt{Observable}. When the observable fires, it calls
\texttt{Observer::notify()}, which submits the thread to the system scheduler
as ready.

\subsection{Observable} trait: \texttt{fn register\_observer(\&self, observer:
Weak<dyn Observer>)}.

\subsection{Observer} trait: \texttt{fn notify(self: Arc<Self>)} --- called by
an Observable when the event occurs. The kernel's \texttt{Waker} implements
\texttt{Observer} by calling \texttt{submit\_ready\_thread(tid)}, queueing the
thread on the least-loaded LP.

\subsection{Wiring} \texttt{sleep(50ms)} creates a \texttt{TimerEvent}
(Observable), calls \texttt{block\_thread(tid, \&event)} to register the
thread's \texttt{Waker} on the event and mark the thread Blocked, pushes the
event into the per-LP timer queue, and yields. When the timer IRQ fires it
processes the event, calls \texttt{signal()} which wakes the thread. The
completion-capability model uses the exact same mechanism: a
\texttt{Completion} is Observable, and \texttt{wait(cap)} blocks on it.

\section{The scheduler}

The system scheduler (\texttt{SYSTEM\_SCHEDULER}) is a per-LP collection of
pluggable \texttt{LpScheduler} instances. The only current implementation is
\texttt{RoundRobin}: threads are queued in a FIFO, and a per-LP quantum timer
(10 ms) marks a context-switch-pending flag.

Key operations:
\begin{itemize}
    \item \texttt{spawn\_thread(asid, entry)} --- kernel-internal constructor
        that creates a Thread, adds it to the master table, and submits it to
        the scheduler. Returns a \texttt{ThreadId}. Userspace thread-spawn
        syscalls do not supply \texttt{asid}; the dispatcher derives it from
        the calling trap frame.
    \item \texttt{block\_thread(tid, \&event)} --- registers the thread's
        \texttt{Waker} on the observable, marks the thread Blocked, removes it
        from the run queue. The thread saves its context on the subsequent
        \texttt{yield\_lp()} and resumes from the same point when woken.
    \item \texttt{submit\_ready\_thread(tid)} --- re-admits a thread to the
        least-loaded LP's run queue. Used by \texttt{Waker::notify()} and other
        wake paths.
    \item \texttt{yield\_lp()} --- sets the pending flag and calls
        \texttt{cond\_yield\_lp()}, which performs the register/stack switch if
        another thread is ready.
\end{itemize}

\texttt{abort()} removes a returning thread from the scheduler and moves it to
\texttt{DEAD\_THREADS} (a deferred reaper list). The reaper drops the thread
from a different thread's context (after the context switch), so the thread's
own kernel stack (which is freed on drop) is no longer in use.

\section{Inter-processor interrupts (IPIs)}

The \texttt{IPI\_CMD\_QUEUES} are per-LP bounded \texttt{ConcurrentQueue}
queues (256 entries by default). Sending a cross-LP work item pushes an RPC to
the target LP's queue and delivers an IPI. The target's IRQ dispatcher drains
the queue in order.

\texttt{IpiRpc} carries kernel-internal RPCs (TLB shootdown, scheduler wakeup)
and a \\ \texttt{Closure(Box<dyn FnOnce() + Send>)} variant for arbitrary
cross-LP work dispatch --- the seed of sitas's typed \texttt{ShardMailbox<M>}.

Two push paths:
\begin{itemize}
    \item \texttt{try\_push\_to} / \texttt{try\_send\_ipi\_rpc} --- returns
        \texttt{Err(rpc)} on full queue (first-class backpressure).
    \item \texttt{push\_to} / \texttt{send\_ipi\_rpc} --- non-fallible
        force-push for must-not-drop kernel RPCs. Evicts the oldest entry if
        full.
\end{itemize}

\section{Capability tables}

A per-address-space \texttt{IdTable<Arc<Completion>>} maps a small integer
\texttt{CompletionCap} (an index into the table) to a kernel object naming an
in-flight or completed operation. The table is bounded (submission
backpressure: \texttt{submit} returns \texttt{WouldBlock} when full).
\texttt{Completion} is \texttt{Observable}: a waiting thread's \texttt{Waker}
is registered on it via \texttt{completion::wait(asid, cap)} inside the
kernel. For real userspace syscalls, that \texttt{asid} is derived from the
running thread's trap frame rather than supplied by userspace.

\endinput
