\chapter{Programming model}

This chapter describes the rules, patterns, and primitives for writing kernel
code that respects the async-first, shard-per-core model. These invariants are
derived from sitas and are enforced by the kernel's own concurrency design.

\section{Invariants}

\subsection{Only the owning LP mutates its state}

Per-LP data is accessed only from the LP that owns it. Cross-LP mutation must
go through message dispatch (\texttt{ShardMailbox<M>}, IPI closure, or IPI
RPC), never through a shared lock acquired from a foreign LP.

\subsection{Messages are owned values}

Everything that crosses a shard boundary is an owned value --- a
\texttt{Box<dyn FnOnce>}, an \texttt{IpiRpc}, or a typed \texttt{M} in a
\texttt{ShardSender<M>}. No shared references, no \texttt{Arc<Mutex<...>>} as
the normal programming model. \texttt{Send + 'static} boundaries enforce this
at compile time.

\subsection{References to shard-local state must not escape}

\texttt{ShardLocal<T>::with(fn)} provides \texttt{\&mut T} inside a closure;
the borrow is verified at runtime (owner-check + re-entrancy flag). The closure
must not store the reference, send it to another thread, or cross an
\texttt{.await}. The same rule applies to sitas's \texttt{ShardLocal<T>}
userspace equivalent.

\subsection{Work stays on its assigned LP}

A thread spawned on LP $N$ stays on LP $N$. To submit work to another LP, use
\texttt{try\_run\_on\_lp(target, closure)} or \texttt{ShardSender::try\_send}.
The kernel's IPI infrastructure and the sitas \texttt{ShardedSubmitter} are
the two sides of this contract.

\subsection{Observability returns owned snapshots}

The kernel's house style is owned values for diagnostics. sitas's
\texttt{RuntimeSnapshot} / \texttt{ShardSnapshot} pattern applies: never
expose borrowed references into runtime internals. Self-tests and the demo
illustrate this.

\section{Primitives}

\subsection{ShardLocal<T>} Lock-free per-LP container. Use for single-LP,
thread-only data. \\Access: \texttt{sl.with(|val| \{ *val = 42; \})}.

\subsection{ShardMailbox<M>} Typed bounded per-LP queue. Cloneable
\texttt{ShardSender<M>} (any LP can send), single-consumer\\
\texttt{ShardReceiver<M>} (one per LP). \texttt{try\_send(m)} returns
\texttt{Err(m)} on backpressure.

\subsection{IPI closure} \texttt{try\_run\_on\_lp(target\_lp, || \{ ... \})}.
Boxes arbitrary work and delivers it to the target LP's IPI queue. Returns the
closure on full queue.

\subsection{Completion API} \texttt{submit}, \texttt{submit\_worker},
\texttt{complete}, \texttt{poll}, \texttt{wait}, \texttt{cancel}, \texttt{close}.
The exit-observer path \\ (\texttt{submit\_worker}) is the intended default; the
explicit \texttt{complete} is the escape hatch for non-thread-based work.

\subsection{CQ ring} \texttt{CompletionQueueRing} for zero-syscall completion
draining. Kernel side: \texttt{write(cap, result)}. Consumer side:
\texttt{read() -> Option<CqEntry>}. Integration: \texttt{open\_as\_with\_cq}
attaches a ring to a capability table; \texttt{complete()} writes to it.

\section{Rule-of-thumb guidance}

\begin{itemize}
    \item \textbf{Use} \texttt{PerLp<T>} for interrupt-accessed data and
        cross-LP access; \textbf{use} \texttt{ShardLocal<T>} for single-LP,
        thread-only data.
    \item \textbf{Use} \texttt{ShardMailbox<M>} (or \texttt{try\_run\_on\_lp})
        for cross-LP communication; do \textbf{not} lock an LP's state from
        another LP.
    \item \textbf{Use} \texttt{submit\_worker} for fire-and-forget async
        kernel work; the worker just returns, and the exit-observer completes
        the capability.
    \item \textbf{Register} a CQ ring when creating an address space so that
        completions are visible to userspace without per-entry syscalls.
    \item \textbf{Accept} backpressure: bounded tables and queues return
        \texttt{WouldBlock} / \texttt{Err(m)}; retry or apply backpressure
        upstream. Never silently drop messages.
\end{itemize}

\section{Working at the sitas--CharlotteOS boundary}

The most useful mental model is that sitas is the userspace runtime and
CharlotteOS is the substrate that makes sitas's assumptions native. sitas code
should think in shards, owned messages, futures, and completion handles.
CharlotteOS code should think in LPs, pinned threads, kernel-owned capability
tables, and wakeable events. The two vocabularies describe the same shape at
different privilege levels.

\subsection{For userspace developers}

Treat a CharlotteOS userspace program as a small \texttt{no\_std} process with
a \texttt{crt0} supplied by \texttt{catten-rt}. The program provides
\texttt{cmain(args, input)} and uses \texttt{catten\_rt::entry!(cmain)}; it
does not define \texttt{\_start}, an allocator, or a panic handler. The type
\texttt{Input<N>} is the declaration of how many bytes the runtime should read
before entering \texttt{cmain}. If the program does not need launch input, use
\texttt{Input<0>}. Command-like values belong in \texttt{Args}.

Userspace application are interested in completion capabilities, mailbox
capabilities, and returned handles.

When writing sitas-on-CharlotteOS code, keep the sitas model intact:
\begin{itemize}
    \item A sitas shard is an LP-affine CharlotteOS user thread.
    \item \texttt{sitas-charlotte::CharlotteReactor::new(lp\_id)} binds a
        reactor to the calling shard's LP.
    \item Async operations return kernel-owned completion capabilities. A
        completion capability is opaque; wait for it, poll it, close it, or
        pass it through a typed channel when the ABI permits that.
    \item Cross-shard work is an owned message. Prefer sitas channels and
        futures over shared mutable state.
    \item Backpressure is normal. Bounded queues and capability tables are part
        of the contract, not exceptional conditions to paper over.
\end{itemize}

The current \texttt{basic\_kv} image is the reference smoke test. It enters via
\texttt{crt0}, uses \texttt{Args([])} and \texttt{Input<0>}, constructs a
Charlotte reactor for LP0, starts a one-shard \texttt{ShardedKv}, writes three
keys, reads one back, and reports \texttt{total\_len == 3} through the
configured output page. If this stops working, either the userspace launch
contract, the syscall ABI, the loader mapping, or the sitas runtime boundary has
regressed.

\subsection{For kernel developers}

The kernel side must preserve the authority and ownership boundaries that make
the userspace model sound:
\begin{itemize}
    \item \textbf{Caller identity is kernel authority.} Syscalls derive the
        caller's address space from \texttt{TrapFrame.asid}; never accept ASID,
        page-table roots, or equivalent authority from userspace registers or
        config pages.
    \item \textbf{Handles are capabilities.} Completion and mailbox ids are
        scoped to the caller's address space. Validate them in the per-AS table
        before acting.
    \item \textbf{Completions must be non-lossy.} A CQ overflow counter is
        useful diagnostics, but a terminal completion must either be delivered,
        retained in a kernel backlog, or prevented by submission backpressure.
    \item \textbf{Buffers have one owner.} While an async operation owns a user
        buffer, userspace must not mutate or free it. Cancellation still ends in
        a terminal completion so ownership can return or be deliberately leaked
        during teardown.
    \item \textbf{Placement is real.} A sitas shard maps to an LP-pinned user
        thread. \texttt{SPAWN\_THREAD} takes an entry virtual address and target
        LP; the address space comes from the caller.
    \item \textbf{Do not smuggle shared kernel state into userspace.} Shared
        pages such as the CQ ring are data queues with a narrow layout, not
        references to kernel objects.
\end{itemize}

The present sitas smoke loader now consumes a stripped AArch64 ELF image rather
than a flattened binary. It maps each \texttt{PT\_LOAD} segment at its linked
virtual address, chooses \texttt{UserCode}, \texttt{UserRoData}, or
\texttt{UserData} from the segment flags, rejects writable-executable segments,
rejects LOAD segments that overlap within one 4 KiB page, and relies on the
freshly zeroed page frames to satisfy the BSS contract. The
\texttt{catten-user} linker script therefore page-separates executable text,
read-only data, and writable data.

This is still a self-test loader, not yet a general userspace process loader:
the config page, CQ page, input page, and heap are still mapped at fixed
CharlotteOS addresses, and there is no dynamic argument/environment layout.
Treat larger sitas binaries as both runtime tests and loader-contract tests.

\endinput
