\chapter{Introduction}

\section{The thesis}

\textbf{Hardware is asynchronous; the operating system and the userspace
runtime should be too.} Two projects, started independently from opposite ends of
the software stack, built toward the same shape and converged on the same
primitive: a waitable completion handle.

\paragraph{CharlotteOS} starts in the kernel and asks: \emph{hardware is
asynchronous with respect to the CPU, so why is the OS synchronous?} Its answer
is to make the kernel async-first --- cheap threads, an observer/waker event
model, and system calls that are asynchronous by default, returning a
completion capability that userspace can wait on.

\paragraph{sitas} starts in userspace and asks: \emph{what does shared-nothing,
shard-per-core concurrency look like when it is built out of Rust ownership
instead of locks?} Its answer is a discipline: each shard owns its state, only
the owning shard mutates it, and everything that crosses a shard boundary is an
owned value moved through an explicit typed message.

Both projects independently converged. sitas's \texttt{Reply<T>} and
\texttt{Notify} (awaitable completions) are the same object as CharlotteOS's
\texttt{Observer} and \texttt{Waker}, just with different names. sitas's shard
mailbox (\texttt{ShardSender}/\texttt{ShardReceiver}) is a bounded per-core
typed queue --- as are CharlotteOS's bounded per-LP \texttt{IPI\_CMD\_QUEUES}.
sitas's \texttt{ShardLocal<T>} (lock-free, owner-checked, closure-based access)
is a Rust-ownership alternative to CharlotteOS's \texttt{PerLp<T>}
(lock-wrapped).

This document describes what happens when these two projects meet: a userspace
runtime whose async model \textbf{agrees} with the kernel it runs on, instead
of emulating async-first on a blocking foundation. That agreement removes the
``retrofit tax'' --- the async-signal-safety, thread pools, and
\texttt{io\_uring}-over-blocking emulation that exist only because the runtime
and the OS disagree about what is fundamental.

\section{What was built and verified}

The collaboration produced a booting, self-testing async-first kernel on both
AArch64 and x86-64, with the following layered architecture:

\textbf{Kernel substrate:}
\begin{itemize}
    \item \textbf{Capability model} --- per-address-space capability tables with
        rights masks.
    \item \textbf{Endpoint IPC} --- Xous-inspired scalar messages, connection
        delegation, memory attachments (Copy/Move/BorrowRead/BorrowWrite),
        deferred replies, reply tokens.
    \item \textbf{Memory objects} --- page-backed ownership transfer between
        protection domains, enforced by the MMU.
    \item \textbf{Completion queues} --- shared-memory ring buffer
        (producer/consumer), non-lossy backlog, per-shard queues, \S8.2 32-byte
        entries, CQ\_WAIT/CQ\_WAKE, CQ\_W AIT\_TIMEOUT.
    \item \textbf{Device capabilities} --- MmioRegion and Interrupt object
        types, user-accessible Device-nGnRnE page mapping, GICv3 SPI routing,
        lock-free interrupt-to-CQ delivery.
\end{itemize}

\textbf{Userspace services:}
\begin{itemize}
    \item \textbf{Service supervision} --- ELF loader, bootstrap capability
        delivery via config page, domain spawn/teardown.
    \item \textbf{Name service} --- human-readable names $\rightarrow$
        re-delegable connections with instance generations; access-key gating.
    \item \textbf{UART driver} --- delegated MMIO + IRQ, direct EL0 MMIO writes,
        interrupt-driven deferred reads, crash/restart reconciliation.
    \item \textbf{Virtio-net driver} --- PCI discovery, BAR0 delegation, virtio
        init sequence, MAC read, virtqueue setup, frame TX/RX (compiled).
\end{itemize}

\textbf{Sitas integration:}
\begin{itemize}
    \item \textbf{ShardExecutor} --- per-shard futures executor with budgeted
        polling and task wakeup from drained events.
    \item \textbf{Per-shard CQ rings} --- loader-contract extension maps one
        ring per shard; targeted wakes prevent cross-shard stealing.
    \item \textbf{ShardParker} --- park/unpark seam over CQ\_WAIT/CQ\_WAKE;
        retire the \texttt{kv::spin\_recv} busy-loop.
    \item \textbf{Waker-integrated channels} --- \texttt{recv().await} registers
        the task waker; \texttt{try\_send/close} invoke it for clean shutdown.
\end{itemize}

\textbf{Networking:}
\begin{itemize}
    \item \textbf{Protocol crates} --- \texttt{charlotte-protocol-net}
        (frame-level), \texttt{charlotte-protocol-msg} (reliable sequenced
        delivery).
    \item \textbf{Smoltcp adapter} --- \texttt{phy::Device} trait over the NIC
        driver endpoint; TCP/IP service binary compiles.
\end{itemize}

\textbf{Syscall ABI:} 49 syscalls across completion, IPC, memory, device, and
CQ operations. Caller ASID is derived from the trap frame, never a userspace
parameter. All verified on QEMU AArch64.

\section{How this document is organized}

\begin{itemize}
    \item \textbf{Chapter 2} --- Architecture: the three primitives, endpoint
        IPC, service supervision, device capabilities, completion queues, sitas
        integration, userspace drivers, networking.
    \item \textbf{Chapter 3} --- Execution model: threads, blocking, context
        switching, timers.
    \item \textbf{Chapter 4} --- The async syscall ABI.
    \item \textbf{Chapter 5} --- Programming model: invariants, ShardLocal,
        ShardMailbox, userspace patterns.
    \item \textbf{Chapter 6} --- Development: build, QEMU boot, self-tests, CI.
    \item \textbf{Chapter 7} --- What we learned: bugs found on hardware, root
        causes, fixes.
    \item \textbf{Chapter 8} --- Current status: phases, success criteria,
        verified features, repositories.
    \item \textbf{Appendix} --- Glossary.
\end{itemize}

\section{Who this is for}

Developers familiar with Rust (ownership, \texttt{Send}, \texttt{Future}),
concurrency primitives, and basic OS architecture. Every claim is based on what
was built and verified on QEMU AArch64 and x86-64. Unimplemented ideas are
flagged.

\endinput
